Future challenges for Research Software Engineering

### 2025-05-18 future_challenges_for_RSEng.tex
update ai section

\subsection{Impact of AI and automation}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
AI is already transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would be impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI accelerates scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
For software development, AI-powered tools automate code generation, bug detection, and testing, significantly increasing productivity and reducing human error.
Developers are shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.
The integration of AI is democratizing access to advanced technologies and analytics, enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster, more efficient, and more innovative than ever before.
With the general availability of AI the interaction with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge,
since AI copilots are available 24/7.
Since AI systems are so democratically available and will be used as black boxes,
the interpretation of their outputs will be far more important.
Currently we are expanding and documenting our work in the open and making our code available,
therefore there will be more structured material to train GenAI on.
However, as people get quicker answers from GenAI, the community resources (\eg Stack Overflow, Wikipedia)
get poorer.

>>>>

\subsection{Impact of AI and automation}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
However, when these tools are used in practice many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and debugging the generated code \cite{Simkute2025}.

The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.

As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster.
With the general availability of AI, the interaction of RSEs with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge.

Another risk of the widespread use of AI tools is that community resources such as \eg Stack Overflow and Wikipedia become poorer because there are fewer incentives to maintain them.
